%==============================================================================
% Final Paper — Emergency Department Wait Time Prediction Using Machine Learning
% IEEE Conference Format
% Course: CSCI-7090 — Data Science and Machine Learning, Georgia Southern Univ.
%==============================================================================
\documentclass[conference]{IEEEtran}
\IEEEoverridecommandlockouts

\usepackage{cite}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{xcolor}
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage{float}
\usepackage{hyperref}
\usepackage{url}

\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
    T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}

\begin{document}

\title{Emergency Department Wait Time Prediction Using\\Machine Learning}

\author{
\IEEEauthorblockN{Vandan Patel}
\IEEEauthorblockA{\textit{Department of Computer Science}\\
\textit{Georgia Southern University}\\
vp04406@georgiasouthern.edu}
\and
\IEEEauthorblockN{Annagrace Howell}
\IEEEauthorblockA{\textit{Department of Computer Science}\\
\textit{Georgia Southern University}\\
ag29516@georgiasouthern.edu}
\and
\IEEEauthorblockN{Yasmin Rocio Orduz Landazabal}
\IEEEauthorblockA{\textit{Department of Computer Science}\\
\textit{Georgia Southern University}\\
yo00553@georgiasouthern.edu}
}

\maketitle

%==============================================================================
\begin{abstract}
Long, unpredictable waits in the Emergency Department (ED) damage patient
outcomes, raise clinical risk, and degrade hospital-wide flow, yet the triage
tools used today rank urgency without telling anyone how long care will
actually take. We address this transparency gap by building and comparing five
regression models---Linear Regression, Ridge, Random Forest, XGBoost, and
LightGBM---that estimate ED length of stay (LOS) in minutes from triage
features alone, using a publicly available CC0 dataset of 9{,}216 ED visits
expanded from 17 raw fields to 26 engineered features. Beyond the standard
overall-accuracy report, we audit the model along three axes that the
literature usually treats in isolation: per-Emergency-Severity-Index (ESI)
performance across all five acuity levels, SHAP-based feature attribution, and
fairness across race/ethnicity, gender, and age group. On a held-out 20\%
test set the regularized linear baseline gave the best aggregate error
(MAE 89.1~min, R$^2$~0.141), narrowly ahead of Random Forest (MAE 91.0) and
ahead of XGBoost (93.8) and LightGBM (93.4); SHAP confirmed that the ESI
acuity score alone carries roughly six times the explanatory weight of any
other feature; ESI-stratified evaluation revealed a near fourfold MAE gap
between Level~5 (37 min) and Level~2 (130 min) patients that the pooled metric
hides; and the fairness audit surfaced a 15.3-minute MAE gap across
race/ethnicity groups and an 11.5-minute gap across age groups, with only a
1.0-minute gap between genders. The work is intended for ED operations leads
who need defensible patient-flow estimates, clinicians who must communicate
realistic timing at triage, equity researchers auditing ML deployments in
healthcare, and patients who deserve a transparent answer to "how long will I
wait?".
\end{abstract}

\begin{IEEEkeywords}
emergency department, wait time prediction, length of stay, machine learning,
regression, triage, ESI, XGBoost, LightGBM, SHAP, algorithmic fairness
\end{IEEEkeywords}

%==============================================================================
\section{Introduction}
%==============================================================================

\subsection{Background and Motivation}
The Emergency Department (ED) is the open door of the modern hospital: it is
where unscheduled, time-sensitive, and clinically heterogeneous patients all
arrive on the same shift and compete for the same finite pool of beds,
nurses, and physicians. Across the United States, Europe, and much of the
Asia-Pacific region, ED visit volumes have outpaced hospital capacity for
more than a decade, and the resulting overcrowding is now classified as a
genuine public-health risk rather than an operational nuisance~\cite{morley2018,
derlet2002}. The downstream effects are well documented. Patients who wait
too long are more likely to suffer adverse events while they wait, more
likely to leave without being seen, and statistically more likely to die in
hospital after admission~\cite{bernstein2009}. From the staff side, sustained
overcrowding drives burnout, increases handoff errors, and pushes ambulance
diversion, which spreads the bottleneck to the surrounding emergency medical
system.

A second motivating thread is operational. Length of stay in the ED has
become one of the most-watched performance indicators in emergency medicine
because it is a single number that condenses everything happening to a
patient between triage and disposition: how long the workup took, whether
the consult arrived, whether an inpatient bed was available, whether the lab
was busy~\cite{ricciardi2024}. When ED LOS becomes unpredictable, a chain of
hospital-wide planning processes degrades with it---bed assignment, transport
scheduling, nurse staffing, even meal delivery. A predictive model that
turns LOS from a guess into a number is therefore valuable not only at the
bedside but at the operations desk.

The third thread is the gap between urgency and timing. The triage scoring
systems EDs already use---primarily the Emergency Severity
Index (ESI) in the United States and the Manchester Triage Scale (MTS) in
Europe~\cite{gilboy2012}---answer the question "who do we see next?" but not
"how long until I am seen?" or "how long until I go home?". These two
questions are clinically and operationally distinct. Machine Learning (ML)
models trained on the same routinely collected triage data---vital signs,
demographics, arrival mode, ESI level---can fill the second question,
producing a continuous patient-level estimate that augments rather than
replaces the existing triage scale. A growing body of recent
work~\cite{wang2025fair, wu2025, wang2025interp} has begun to populate this
space, but most published efforts still cast the problem as a binary
"prolonged vs. not prolonged" classification, train on proprietary
single-center data, or report a single pooled accuracy metric that hides
performance differences between patient subgroups.

\subsection{Research Questions}
The problem we address is the absence, in current practice, of a transparent
patient-level estimate of how long an ED visit will take. The estimate must be
computable at the moment of triage, must use only data already collected
during the standard triage interaction, must come with an explanation that
clinical staff can interpret, and must behave consistently across patient
subgroups---a model that is twice as wrong for one demographic group as for
another is not deployable in a clinical setting regardless of its average
accuracy. We split the problem into two research questions, each chosen to
target a different facet of trustworthiness:

\begin{itemize}
    \item \textbf{RQ1 (Predictive accuracy and generalizability across triage
    levels).} Can supervised regression models trained on routinely collected
    ED triage data predict patient LOS in minutes with materially lower error
    than naive baselines, and does that accuracy generalize across all five
    ESI triage levels, or is it driven by a single dominant sub-population?
    \item \textbf{RQ2 (Explainability, interpretability, and demographic
    fairness).} Which features actually drive the model's LOS predictions
    according to SHAP attribution, do those drivers align with clinical
    intuition and with the patterns reported in prior interpretable-ML
    studies, and does the resulting model produce equitable error across
    subgroups defined by race/ethnicity, gender, and age?
\end{itemize}

\subsection{Objectives}
The research questions translate to four concrete project objectives:
\begin{enumerate}
    \item Train and benchmark five regression models---Linear Regression,
    Ridge, Random Forest, XGBoost, and LightGBM---on an 80/20 hold-out split
    with 5-fold cross-validation on the training fold, reporting MAE, RMSE,
    and R$^2$.
    \item Refit the best model separately within each of the five ESI levels
    and compare per-level error, exposing the subgroup variation that pooled
    metrics hide.
    \item Apply SHAP TreeExplainer to the best gradient-boosted model and
    report a global feature ranking, both for clinical sanity-checking and
    for direct comparison with previously published SHAP rankings.
    \item Disaggregate the test-set MAE and RMSE across race/ethnicity,
    gender, and age group, and quantify the worst-case subgroup gap.
\end{enumerate}

\subsection{Contributions}
Against the existing literature, this paper offers four contributions:
\begin{enumerate}
    \item \textbf{A combined methodology.} Our pipeline simultaneously
    delivers continuous LOS regression, ESI-stratified evaluation, SHAP
    interpretability, and demographic fairness auditing on a single
    publicly accessible dataset. None of the twelve studies surveyed in
    Section~\ref{sec:lit} delivers all four together.
    \item \textbf{An empirical counter-finding.} On a moderately sized
    dataset, regularized linear regression matches or slightly beats the
    gradient-boosted models, in tension with the prevailing claim that
    XGBoost and LightGBM dominate ED prediction across the
    board~\cite{wu2025, ricciardi2024}. We argue this is a dataset-size
    effect worth reporting honestly rather than tuning around.
    \item \textbf{An equity finding.} A non-trivial 15.3-minute MAE gap
    across race/ethnicity groups exists in the model's errors and is
    invisible if only a single pooled metric is reported. This argues for
    subgroup auditing as a default, not an optional extension, in ED ML.
    \item \textbf{A reproducibility artifact.} A cleaned 26-feature dataset,
    a documented analysis notebook, and saved per-model and per-subgroup
    metrics are released alongside this paper so that other teams can
    replicate, criticize, or extend the experiment.
\end{enumerate}

\subsection{Paper Roadmap}
Section~\ref{sec:lit} reviews twelve recent peer-reviewed studies on ML for
ED triage and patient-flow prediction, ending with an explicit list of
recurring gaps and a positioning paragraph that locates our contribution
inside the literature. Section~\ref{sec:methods} covers the dataset, a
condensed preprocessing pipeline, the five candidate models, and the
evaluation protocol. Section~\ref{sec:results} reports results in four
layers---overall, ESI-stratified, SHAP-based, and fairness-disaggregated---
and compares each layer against the closest prior-work benchmark.
Section~\ref{sec:discussion} interprets the results, lists strengths and
limitations honestly, and pulls out practical implications.
Section~\ref{sec:conclusion} concludes and proposes future work. All
numerical tables are collected in the Appendix and referenced inline by
number throughout.

%==============================================================================
\section{Literature Review}\label{sec:lit}
%==============================================================================

\subsection{Theoretical Background}
Three research traditions converge on the ED prediction problem. The first
is clinical triage, which since the late 1990s has been formalized through
ordinal scales such as ESI~\cite{gilboy2012} that translate vital signs and
chief complaint into a 1--5 priority. The second is the rise of
gradient-boosted decision trees as the default tabular learner in
healthcare~\cite{chen2016xgboost, ke2017lightgbm, breiman2001}; on most
clinical structured-data problems they sit at the Pareto front of accuracy
and training cost, which is why they dominate the recent ED literature. The
third is post-hoc explainability through SHAP~\cite{shap, lundberg2018},
which lets a tree ensemble be interrogated at the feature level without
sacrificing predictive accuracy. The standard recipe in the recent ED
literature is to combine all three: feed triage features into a boosted
ensemble, evaluate on AUC for classification or on MAE/RMSE/R$^2$ for
regression~\cite{wu2025, vural2025}, and explain the result with SHAP. Our
study follows this recipe and tests where it breaks.

\subsection{Synthesis of the Twelve Reviewed Studies}
We organize the twelve studies into five themes: systematic reviews,
empirical wait-time and LOS models, benchmark and interpretability work,
fairness and overcrowding studies, and advanced AI paradigms. The full
synthesis appears as Table~\ref{tab:summary} in
Appendix~\ref{app:tables}; the reading of the table that follows is the
narrative the table is designed to support.

\subsection{Systematic Reviews}
Two systematic reviews anchor the field. Porto~\cite{porto2024} surveys ML
and natural-language-processing approaches to ED triage in a 2024
PRISMA-2020 review and concludes that learned models consistently outperform
rule-based triage scales, while flagging the absence of continuous-wait-time
regression and the limited use of explainability tools as the two largest
unmet needs. The second 2025 systematic review~\cite{slr2025} surveys ML for
predicting inpatient admission from ED triage and reports a tight AUC range
of 0.81--0.93 across rigorous studies, with random forest and gradient
boosting at the top of the leaderboard.

\subsection{Empirical Wait-Time and LOS Models}
Five empirical studies form the most directly comparable group to our work.
Wang et al.~\cite{wang2025fair} train an interpretable XGBoost classifier on
prolonged ($\geq$30 min) wait time in ESI-3 patients, achieving AUC $\approx$
0.72 and documenting performance gaps across race/ethnicity and insurance
type, but remain restricted to the binary target and a single ESI bucket.
Ricciardi et al.~\cite{ricciardi2024} apply random forest, decision tree, and
SVM to nearly half a million Italian ED admissions and reach AUC $\approx$
0.80 for the binary prolonged-LOS target, validating that triage features
alone carry useful signal. Seo et al.~\cite{seo2024} extend the feature space
with NLP-processed chief-complaint text and find LightGBM the best of their
suite. Wang, Sambamoorthi et al.~\cite{wang2025interp} compare five learners
on the same prolonged-wait-time target as~\cite{wang2025fair} and confirm
XGBoost/RF as the strongest. Wu et al.~\cite{wu2025} stand out as the
clearest precedent for our setup: they treat ED LOS as a continuous
regression target on 187{,}028 records, identify LightGBM as the strongest
model, and surface "current emergency waiting time"---a real-time queue
signal---as the dominant feature.

\subsection{Benchmarks and Interpretability}
Two studies build infrastructure rather than fit models on their own. Xie
et al.~\cite{xie2022} release the MIMIC-IV-ED benchmark and a reference
preprocessing pipeline for the wider research community, demonstrating that
gradient boosting and random forest beat traditional clinical scoring
systems. Sulaiman et al.~\cite{sulaiman2025} push the explainability
frontier past SHAP by combining ensemble ML with TE2Rules rule extraction on
ten minimal features, sketching a path toward clinician-readable
"if-then" explanations.

\subsection{Fairness, Overcrowding, and Advanced AI}
The remaining three studies cover the wider context. Vural
et al.~\cite{vural2025} target system-level overcrowding rather than
individual-patient prediction, using time-series transformers (TSiTPlus,
XCMPlus) on 6-hour and 24-hour occupancy and reporting MAE $\approx$ 4.19
patients with R$^2$ $\approx$ 0.56. Lett et al.~\cite{lett2025} compare
intersectional and marginal fairness debiasing for ED admission prediction
on MIMIC-IV-ED; both methods preserve overall AUROC at 0.85, but the
intersectional approach roughly doubles the calibration improvement (22.3\%
vs.~11.3\% subgroup error reduction). Sun et al.~\cite{sun2024} introduce
ED-Copilot, which pairs BioGPT with a reinforcement-learning agent to
recommend lab orderings that minimize wait time and demonstrate a roughly
50\% reduction in average laboratory wait time---the most ambitious AI
paradigm in the review.

\subsection{Identified Research Gaps}
Reading Table~\ref{tab:summary} together with the per-study accounts above
makes five gaps visible. \textit{Gap 1: continuous regression is rare}---ten
of the twelve studies treat their target as a binary class; only Wu et
al.~\cite{wu2025} and Seo et al.~\cite{seo2024} model wait time or LOS as a
continuous variable in minutes. \textit{Gap 2: triage segmentation is
absent}---no study reports performance broken down across all five ESI
levels; subgroup analyses, when they exist, focus on ESI-3 alone.
\textit{Gap 3: explainability is uneven}---only four of twelve studies apply
SHAP, only one~\cite{sulaiman2025} explores rule-based alternatives, and
seven publish no interpretability analysis at all. \textit{Gap 4:
reproducibility is limited}---nine of twelve studies use proprietary
single-center data, which prevents direct replication. \textit{Gap 5:
fairness audits are scarce}---only two studies~\cite{wang2025fair, lett2025}
formally audit subgroup performance, and both restrict their audit to
classification rather than continuous regression.

\subsection{Positioning of This Study}
Our work is designed to attack all five gaps at once on a single
publicly-released dataset. We model LOS as a continuous variable in minutes
(addresses Gap 1), evaluate every model across all five ESI levels rather
than collapsing to a binary cut (Gap 2), apply SHAP TreeExplainer to our
best gradient-boosted model and publish the global ranking (Gap 3), use the
CC0 Hospital Emergency Room Visits dataset and release a cleaned 26-feature
derivative (Gap 4), and disaggregate test-set MAE and RMSE across
race/ethnicity, gender, and age group (Gap 5). To our knowledge no prior ED
study delivers all five.

%==============================================================================
\section{Methodology}\label{sec:methods}
%==============================================================================

\subsection{Dataset}
The data come from the Hospital Emergency Room Visits dataset on Kaggle,
released under a CC0 (public-domain) license, with a feature schema designed
to mirror the openly published MIMIC-IV-ED benchmark~\cite{xie2022, mimic}.
The raw file contains 9{,}216 ED visits and 17 fields covering structured
triage data, vitals, demographics, arrival context, and disposition.
Table~\ref{tab:dictionary} (Appendix) summarizes every raw field grouped by
clinical role: identifier, demographics, arrival context, triage acuity,
vital signs, and disposition. The target field is
\texttt{ed\_los\_minutes}, the total minutes a patient spends in the
department from arrival to physical departure. As Table~\ref{tab:target}
(Appendix) shows, the empirical distribution of this variable is heavily
right-skewed---a mean of 189.8 min sitting well above the median of 153.7
min, and a long tail reaching 2{,}644.9 min for boarding cases---which is
typical of ED stay-time data and motivates a non-Gaussian modeling
treatment.

\subsection{Data Cleaning and Preprocessing}
Preprocessing was applied as a single sequential pipeline. Numeric vital
signs with up to 12\% missingness were imputed by their median; the four
categorical fields with sparse missing entries were imputed by their mode;
clinical reference bounds (HR 30--250 bpm, SBP 60--280 mmHg, LOS
5--4{,}320 min, etc.) were used to confirm no implausible values survived
imputation; categorical features were label-encoded; five new features were
engineered from existing ones (\textit{shock\_index} = HR/SBP,
\textit{pulse\_pressure} = SBP$-$DBP, three-bucket \textit{arrival\_shift},
binary \textit{is\_weekend}, four-bucket \textit{age\_group}); and finally
all 19 numeric model features were standardized to zero mean and unit
variance, in line with the convention adopted by Wu et al.~\cite{wu2025}
and Wang et al.~\cite{wang2025fair}. Table~\ref{tab:wrangling} (Appendix)
gives the row-by-row before/after summary, ending with 9{,}216 records,
26 features, and zero remaining missing cells; Table~\ref{tab:missing}
profiles the missingness as it appeared on entry to the pipeline.

\subsection{Model Feature Set}
After preprocessing, the regression input vector consists of 19 standardized
features grouped by source: \textit{patient demographics} (age,
gender\_encoded, race\_encoded, age\_group\_encoded), \textit{vital signs}
(heartrate, sbp, dbp, temperature, o2sat, resprate), \textit{triage
information} (acuity, pain\_score), \textit{engineered physiology}
(shock\_index, pulse\_pressure), \textit{arrival context} (arrival\_hour,
is\_weekend, arrival\_transport\_encoded, arrival\_shift\_encoded), and
\textit{visit-level disposition} (was\_admitted). Pearson screening
confirmed that only \texttt{acuity} has a non-trivial linear correlation
with the target ($r=-0.38$); every other feature falls in $|r|\leq 0.02$,
which is the empirical signature of a problem dominated by one strong driver
plus weak nonlinear and interaction signal in the remainder.

\subsection{Models and Why We Chose Them}
We evaluated five regressors covering three families.
\textit{Linear baselines}: Linear Regression and Ridge ($\alpha=1.0$); these
establish a reference floor and check whether L2-regularization helps
under the multicollinearity present between SBP, DBP, and pulse\_pressure.
\textit{Tree baseline}: Random Forest with 200 trees, max depth 14, and
minimum 4 samples per leaf, following the strong RF results reported
by~\cite{ricciardi2024} and the bagging argument of~\cite{breiman2001}.
\textit{Gradient boosting}: XGBoost~\cite{chen2016xgboost} with 400 trees,
learning rate 0.05, max depth 6, and 0.85 subsample/colsample fractions; and
LightGBM~\cite{ke2017lightgbm} with 500 trees, 31 leaves, learning rate
0.05. The boosting hyperparameters were chosen by light manual tuning on
the training fold to match the configurations most commonly reported in the
ED ML literature~\cite{wu2025, seo2024, wang2025fair} so that any difference
in our findings cannot be charged to underpowered baselines. All models use
\texttt{random\_state=42}.

\subsection{Evaluation Protocol}
The 9{,}216 records were split 80/20 into a training set (7{,}372 records)
and a held-out test set (1{,}844 records) using a fixed seed. We report
three metrics: MAE (interpretable in clinical units---minutes of error),
RMSE (which over-weights large errors and so penalizes the long-tail
boarding cases), and R$^2$ (proportion of LOS variance explained). All three
are reported both on the held-out test set and as a 5-fold cross-validation
average on the training fold. For ESI-stratified evaluation, we re-split the
data within each acuity level before refitting Ridge to keep training- and
test-fold separation. For SHAP analysis we apply
TreeExplainer~\cite{shap, lundberg2018} to the LightGBM model and report
mean $|$SHAP$|$ per feature on the test set. For the fairness audit, we
compute MAE and RMSE separately for each group of the
race/ethnicity, gender, and age-group categorical variables on the same
held-out test set. The full software stack is Python 3.11 with
scikit-learn 1.8.0, XGBoost 3.2.0, LightGBM 4.6.0, and SHAP 0.51.0; results
files and the analysis notebook are released with the paper.

%==============================================================================
\section{Results}\label{sec:results}
%==============================================================================

\subsection{Overall Model Performance}
Table~\ref{tab:overall} (Appendix) reports test-set and 5-fold CV results
for all five models in a single synthesis. Ridge Regression yielded the
lowest test MAE at 89.09 min and the highest test R$^2$ at 0.141, edging
out plain Linear Regression by less than 0.01 on every metric (the two
collapse to nearly the same model when the regularization weight is small
relative to the gradient of the loss surface). Random Forest came next at
MAE 91.03; XGBoost and LightGBM trailed at 93.80 and 93.42. Cross-validation
metrics tracked the held-out test ranking to within $\pm$1 MAE point for
every model, ruling out overfitting as an explanation. As a sanity floor,
the naive "predict the training-set mean" baseline produced a test MAE of
98.71, so the weakest of our five models still recovers around 5 minutes of
error reduction over guessing the average.

\textit{Comparison with prior work.} Wu et al.~\cite{wu2025}, the most
directly comparable continuous-regression study, achieve LightGBM R$^2$
between roughly 0.40 and 0.55 on 187{,}028 records, with the bulk of their
predictive lift coming from a real-time "current emergency waiting time"
queue-depth feature that is unavailable in our static dataset.
Ricciardi et al.~\cite{ricciardi2024} report AUC $\approx$ 0.80 on a binary
prolonged-LOS target on roughly 500{,}000 records---a different metric on a
different framing, but a useful upper anchor for what triage features alone
can deliver. The most striking finding in our table, given the literature
backdrop, is that on this dataset \emph{regularized linear regression
matches the best non-linear model}. The straightforward explanation,
consistent with the correlation profile of the features ($r=-0.38$ for
acuity, $|r|\leq 0.02$ for everything else), is that this prediction problem
contains one strong linear driver and very little exploitable interaction
signal at the available sample size. Boosting needs interactions to pay for
its capacity; on a dataset of 9{,}216 records with one dominant linear
effect, regularized linear models can match it.

\subsection{ESI-Stratified Performance}
Refitting Ridge separately within each acuity level produced
Table~\ref{tab:esi} (Appendix). Two patterns emerge that the pooled metric
hides. First, MAE scales almost linearly with mean LOS by triage level:
ESI-5 patients (mean LOS 91 min, lowest acuity) have MAE 36.96 min, while
ESI-2 patients (mean LOS 265 min, second-highest acuity) have MAE 130.28
min---a 3.5$\times$ ratio that the single pooled MAE of 89 min would never
reveal. Second, every per-level R$^2$ value sits between $-0.16$ and $-0.01$,
even though the pooled R$^2$ is 0.141. This is not a contradiction. The
pooled R$^2$ benefits from the strong between-level variance driven by
acuity itself; once that variance is removed by stratifying, the residual
within-level signal in vital signs and demographics is too weak for the
linear model to capture. The clinical implication is that a pooled-metric
deployment would systematically over-promise accuracy for high-acuity
patients and under-promise it for low-acuity patients. We did not find this
pattern reported in any of the twelve reviewed studies.

\subsection{SHAP Feature Importance}
Table~\ref{tab:shap} (Appendix) reports global mean $|$SHAP$|$ values for
the LightGBM model on the held-out test set, ranked from largest to
smallest. The ESI \texttt{acuity} score dominates at 46.07---approximately
6.4$\times$ the second feature. The next tier consists of physiology:
\texttt{o2sat} (7.15), \texttt{dbp} (6.27), \texttt{temperature} (5.88),
\texttt{age} (5.87), \texttt{pulse\_pressure} (5.85), \texttt{resprate}
(5.20), \texttt{shock\_index} (5.07), \texttt{heartrate} (4.96),
\texttt{sbp} (3.94). Demographics, arrival context, and visit-level
disposition---features that a fairness-skeptical reader might worry the
model leans on---occupy the bottom half of the ranking, with
\texttt{age\_group\_encoded} dead last at 0.62.

\textit{Comparison with prior work.} The dominance of the acuity score
matches Wang et al.~\cite{wang2025fair}, whose XGBoost SHAP analysis put ESI
level at the top of the list, and Wang, Sambamoorthi
et al.~\cite{wang2025interp}, who placed ESI level and arrival hour as
their top two. Our \texttt{arrival\_hour} ranks 13th rather than near the
top, which we attribute to two differences: the prior studies' EDs had
strong queue-depth dynamics that load onto arrival hour as a proxy, and our
dataset has a more uniform arrival-hour distribution that flattens this
proxy effect. The high SHAP rank of vital signs over administrative and
demographic features matches the clinically intuitive story that
physiological severity drives LOS more than identity does.

\subsection{Demographic Fairness Audit}
Tables \ref{tab:fair_race}, \ref{tab:fair_gender}, and \ref{tab:fair_age}
(all in Appendix) report MAE and RMSE for the Ridge model disaggregated
across race/ethnicity, gender, and age group, respectively. The headline
numbers are a 15.26-minute MAE gap between the worst-served group
(Black/African American, MAE 91.23) and the best-served group (OTHER, MAE
75.97); a 1.03-minute MAE gap between female (88.61) and male (89.64)
patients; and an 11.50-minute MAE gap between the worst-served age group
(Pediatric, MAE 95.60) and the best-served (Middle Age, MAE 84.10). The
smaller demographic groups---HISPANIC, ASIAN, OTHER---have lower errors than
the two large groups (WHITE and Black), which most likely reflects
narrower within-group LOS variance rather than a genuinely better model fit
for those subgroups; we discuss this caveat in Section~\ref{sec:discussion}.

\textit{Comparison with prior work.} Wang et al.~\cite{wang2025fair}
report comparable relative AUC gaps across race and insurance in their
classification setup; ours is the same phenomenon expressed in regression
units. Lett et al.~\cite{lett2025} demonstrate a 22.3\% subgroup
calibration-error reduction from intersectional debiasing on a similar ED
prediction task, suggesting a concrete mitigation path our descriptive
audit identifies but does not yet apply. The much smaller gender gap and
the larger race/age gaps echo a well-documented pattern in the wider
clinical-ML fairness literature~\cite{rajkomar2018, obermeyer2019} that
binary gender is a less common source of disparity than race or age in
standard structured-data models.

%==============================================================================
\section{Discussion}\label{sec:discussion}
%==============================================================================

\subsection{What the Results Mean}
Three findings deserve to be foregrounded. First, the linear-vs-boosted
ordering on this dataset is a counter-finding to the prevailing
literature pattern that XGBoost and LightGBM dominate ED prediction. Read
together with the SHAP result that one feature carries roughly half the
total predictive signal, and with the very weak linear correlations of every
other feature, the parsimonious explanation is that on a moderately sized
dataset with one dominant linear driver and a weakly informative remainder,
regularized linear regression has nothing to lose and boosting has too
little interaction signal to gain. We expect the ordering to flip on much
larger datasets or on datasets enriched with real-time queue-depth signals.
Second, ESI stratification reveals that the single pooled metric overstates
how well the model serves high-acuity patients and understates how well it
serves low-acuity patients; reporting a single MAE without stratification
is therefore quietly misleading for any clinical deployment. Third, the
race/ethnicity MAE gap of 15.26 minutes is large enough to matter for
patient communication---the difference between "tell me 90 minutes" and
"tell me 105 minutes" is the difference between trust and distrust the next
time the patient walks in.

\subsection{Strengths}
The study has four strengths worth naming. (1) The dataset is fully
public under a permissive license, the random seed is fixed, and the
cleaned CSV plus analysis notebook are released with the paper, so the
experiment is reproducible end-to-end. (2) The pipeline simultaneously
delivers continuous regression, ESI stratification, SHAP, and a fairness
audit---a combination not present in any of the twelve reviewed studies.
(3) We include simple regularized-linear baselines and report honestly
that they win, against the publication-bias gradient toward boosting in
this literature. (4) Every demographic subgroup with at least 50 test-set
records is reported.

\subsection{Limitations}
We acknowledge five limitations. \textit{Sample size.} 9{,}216 visits is
modest compared with the 187{,}028 records of Wu et al.~\cite{wu2025} or
the $\sim$400{,}000 of MIMIC-IV-ED~\cite{xie2022}; both the linear-vs-
boosted ordering and the per-subgroup error estimates carry meaningful
uncertainty at our sample size. \textit{Single source.} The dataset comes
from a single ED, so external validation on a different institution is
required before the model can be claimed to generalize. \textit{No queue
signal.} The static feature vector cannot include real-time crowding
indicators (current census, mean wait of patients ahead in queue, board
status); Wu et al.~\cite{wu2025} report that exactly this feature is the
strongest predictor in their model, and its absence in our setup likely
caps the achievable R$^2$. \textit{Limited explanatory power.} The pooled
R$^2$ of 0.141 means 86\% of the variance in LOS is unexplained by triage
features alone, motivating future incorporation of NLP on chief-complaint
text~\cite{seo2024} and laboratory turnaround features. \textit{Descriptive
fairness only.} We report subgroup gaps but do not yet apply the
intersectional-debiasing correction of Lett et al.~\cite{lett2025}; that
mitigation is left for follow-on work.

\subsection{Practical Implications}
For ED operations leads, these results suggest that a Ridge model deployed
at triage on this kind of feature set can deliver a patient-level estimate
accurate to roughly $\pm$90 minutes on average, with substantially better
accuracy ($\pm$37 min) for the lowest-acuity patients and substantially
worse accuracy ($\pm$130 min) for the second-highest acuity tier. Patient
communication should respect this stratification: a confident "30 to 60
minutes" is realistic for ESI-5 patients but not for ESI-2 patients, who
should be told something like "two to six hours, depending on workup." For
equity researchers, the 15-minute race-related MAE gap is a concrete and
reproducible target for a debiasing intervention. For clinicians, the SHAP
ranking is a useful sanity check: the model is leaning on physiological
severity, not on demographics or administrative features.

%==============================================================================
\section{Conclusion and Future Work}\label{sec:conclusion}
%==============================================================================

We presented an end-to-end regression pipeline for predicting Emergency
Department length of stay as a continuous variable, evaluated five models on
9{,}216 ED visits with an 80/20 holdout and 5-fold cross-validation,
stratified results across all five ESI levels, applied SHAP TreeExplainer
to interpret the best gradient-boosted model, and audited demographic
fairness across race/ethnicity, gender, and age group. Ridge Regression
produced the lowest test MAE (89.09 min, R$^2$~$=0.141$), narrowly ahead of
Random Forest and ahead of the two gradient-boosted models---a result that
nuances the literature's general preference for XGBoost and LightGBM in this
domain. SHAP confirmed that ESI acuity dominates the prediction signal. ESI
stratification revealed an MAE gradient from 37 (ESI 5) to 130 (ESI 2)
minutes that is invisible in pooled metrics. The fairness audit surfaced a
15.26-minute MAE gap across race/ethnicity and an 11.50-minute gap across
age groups, with a near-zero gap across genders.

Three lines of follow-on work are most promising. First, incorporating
NLP-processed chief-complaint text along the lines of Seo et
al.~\cite{seo2024} should add semantic signal that the static-feature
formulation cannot capture. Second, replicating the experiment on
MIMIC-IV-ED~\cite{xie2022, mimic} and on a second institutional dataset
would test external validity. Third, applying the intersectional
debiasing of Lett et al.~\cite{lett2025} would translate the descriptive
fairness gaps reported here into measurable mitigation outcomes. We also
plan to layer rule extraction in the style of Sulaiman
et al.~\cite{sulaiman2025} on top of SHAP to produce clinician-facing
"if--then" explanations alongside the feature-attribution view.

%==============================================================================
\section*{Acknowledgments}
%==============================================================================
This work was conducted as part of the CSCI-7090 Data Science and Machine
Learning course project, Team~3, Spring 2026, Georgia Southern University.
We thank Prof.~Vijayalakshmi Ramasamy for guidance on framing and
structure.

%==============================================================================
\begin{thebibliography}{00}

\bibitem{porto2024}
B. M. Porto, ``Improving triage performance in emergency departments using
machine learning and natural language processing: a systematic review,''
\textit{BMC Emerg. Med.}, vol. 24, no. 1, p. 219, 2024. [Online]. Available:
\url{https://doi.org/10.1186/s12873-024-01135-2}

\bibitem{slr2025}
Multiple authors, ``Predicting inpatient admissions from emergency department
triage using machine learning: a systematic review,'' \textit{JACEP Open},
Elsevier, 2025. [Online]. Available:
\url{https://www.sciencedirect.com/science/article/pii/S2949761225000045}

\bibitem{wang2025fair}
H. Wang et al., ``Evaluating fairness of machine learning prediction of
prolonged wait times in emergency department with interpretable eXtreme
gradient boosting,'' \textit{PLOS Digit. Health}, vol. 4, no. 3, 2025.
[Online]. Available: \url{https://doi.org/10.1371/journal.pdig.0000751}

\bibitem{xie2022}
F. Xie, J. Zhou, J. W. Lee et al., ``Benchmarking emergency department
prediction models with machine learning and public electronic health
records,'' \textit{Sci. Data}, vol. 9, p. 658, Oct. 2022. [Online].
Available: \url{https://doi.org/10.1038/s41597-022-01782-9}

\bibitem{ricciardi2024}
C. Ricciardi et al., ``Evaluation of different machine learning algorithms
for predicting the length of stay in the emergency departments: a
single-centre study,'' \textit{Front. Digit. Health}, vol. 5, p. 1323849,
Jan. 2024. [Online]. Available:
\url{https://doi.org/10.3389/fdgth.2023.1323849}

\bibitem{seo2024}
H. Seo et al., ``Prediction of hospitalization and waiting time within 24
hours of emergency department patients with unstructured text data,''
\textit{Health Care Manag. Sci.}, vol. 27, pp. 114--129, 2024. [Online].
Available: \url{https://doi.org/10.1007/s10729-023-09660-5}

\bibitem{wang2025interp}
H. Wang, N. Sambamoorthi, D. Sandlin, and U. Sambamoorthi, ``Interpretable
machine learning models for prolonged emergency department wait time
prediction,'' \textit{BMC Health Serv. Res.}, vol. 25, p. 403, Mar. 2025.
[Online]. Available: \url{https://doi.org/10.1186/s12913-025-12535-w}

\bibitem{sulaiman2025}
W. A. Sulaiman et al., ``Leveraging machine learning and rule extraction for
enhanced transparency in emergency department length of stay prediction,''
\textit{Front. Digit. Health}, vol. 6, p. 1498939, Feb. 2025. [Online].
Available: \url{https://doi.org/10.3389/fdgth.2024.1498939}

\bibitem{wu2025}
W. Wu et al., ``Development of an emergency department length-of-stay
prediction model based on machine learning,'' \textit{World J. Emerg. Med.},
vol. 16, no. 3, pp. 220--224, 2025. [Online]. Available:
\url{https://doi.org/10.5847/wjem.j.1920-8642.2025.048}

\bibitem{vural2025}
O. Vural et al., ``An artificial intelligence-based framework for predicting
emergency department overcrowding: development and evaluation study,''
\textit{JMIR Med. Inform.}, vol. 13, p. e73960, 2025. [Online]. Available:
\url{https://doi.org/10.2196/73960}

\bibitem{lett2025}
E. Lett et al., ``Intersectional and marginal debiasing in prediction models
for emergency admissions,'' \textit{JAMA Netw. Open}, vol. 8, no. 5,
p. e2512947, May 2025. [Online]. Available:
\url{https://doi.org/10.1001/jamanetworkopen.2025.12947}

\bibitem{sun2024}
L. Sun et al., ``ED-Copilot: reduce emergency department wait time with
language model diagnostic assistance,'' in \textit{Proc. 41st Int. Conf.
Mach. Learn. (ICML)}, PMLR, vol. 235, pp. 46942--46956, Jul. 2024. [Online].
Available: \url{https://proceedings.mlr.press/v235/sun24a.html}

\bibitem{mimic}
A. Johnson et al., ``MIMIC-IV-ED (Version 2.2),'' PhysioNet, Jan. 2023.
[Online]. Available: \url{https://doi.org/10.13026/5ntk-km72}

\bibitem{shap}
S. M. Lundberg and S.-I. Lee, ``A unified approach to interpreting model
predictions,'' in \textit{Adv. Neural Inf. Process. Syst.}, vol. 30, 2017.

% ---- 5 NEW REFERENCES ADDED FOR THE FINAL PAPER ----

\bibitem{morley2018}
C. Morley, M. Unwin, G. M. Peterson, J. Stankovich, and L. Kinsman,
``Emergency department crowding: A systematic review of causes, consequences
and solutions,'' \textit{PLOS ONE}, vol. 13, no. 8, p. e0203316, Aug. 2018.
[Online]. Available: \url{https://doi.org/10.1371/journal.pone.0203316}

\bibitem{gilboy2012}
N. Gilboy, P. Tanabe, D. Travers, and A. M. Rosenau, ``Emergency Severity
Index (ESI): A Triage Tool for Emergency Department Care, Version 4.
Implementation Handbook,'' Agency for Healthcare Research and Quality,
Rockville, MD, AHRQ Publication No. 12-0014, 2012.

\bibitem{breiman2001}
L. Breiman, ``Random forests,'' \textit{Machine Learning}, vol. 45, no. 1,
pp. 5--32, Oct. 2001. [Online]. Available:
\url{https://doi.org/10.1023/A:1010933404324}

\bibitem{chen2016xgboost}
T. Chen and C. Guestrin, ``XGBoost: A scalable tree boosting system,'' in
\textit{Proc. 22nd ACM SIGKDD Int. Conf. Knowl. Discov. Data Min.}, 2016,
pp. 785--794. [Online]. Available:
\url{https://doi.org/10.1145/2939672.2939785}

\bibitem{ke2017lightgbm}
G. Ke et al., ``LightGBM: A highly efficient gradient boosting decision
tree,'' in \textit{Adv. Neural Inf. Process. Syst.}, vol. 30, 2017,
pp. 3146--3154.

% ---- additional supporting references used in the body ----

\bibitem{lundberg2018}
S. M. Lundberg, G. G. Erion, and S.-I. Lee, ``Consistent individualized
feature attribution for tree ensembles,'' \textit{arXiv:1802.03888}, 2018.
[Online]. Available: \url{https://arxiv.org/abs/1802.03888}

\bibitem{rajkomar2018}
A. Rajkomar, M. Hardt, M. D. Howell, G. Corrado, and M. H. Chin, ``Ensuring
fairness in machine learning to advance health equity,'' \textit{Annals of
Internal Medicine}, vol. 169, no. 12, pp. 866--872, Dec. 2018. [Online].
Available: \url{https://doi.org/10.7326/M18-1990}

\bibitem{obermeyer2019}
Z. Obermeyer, B. Powers, C. Vogeli, and S. Mullainathan, ``Dissecting racial
bias in an algorithm used to manage the health of populations,''
\textit{Science}, vol. 366, no. 6464, pp. 447--453, Oct. 2019. [Online].
Available: \url{https://doi.org/10.1126/science.aax2342}

\bibitem{derlet2002}
R. W. Derlet and J. R. Richards, ``Emergency department overcrowding in
Florida, New York, and Texas,'' \textit{Southern Medical Journal}, vol. 95,
no. 8, pp. 846--849, 2002.

\bibitem{bernstein2009}
S. L. Bernstein et al., ``The effect of emergency department crowding on
clinically oriented outcomes,'' \textit{Acad. Emerg. Med.}, vol. 16, no. 1,
pp. 1--10, Jan. 2009. [Online]. Available:
\url{https://doi.org/10.1111/j.1553-2712.2008.00295.x}

\end{thebibliography}

\clearpage

%==============================================================================
\appendix
%==============================================================================
This appendix collects every numerical table referenced in the body of the
paper. Each table is structured as a synthesis grouped by theme, method, or
clinical role---rather than a flat list---so that the reader can scan for
patterns across rows.
\label{app:tables}

% Tighten spacing between consecutive tables in the appendix
\setlength{\textfloatsep}{6pt plus 1pt minus 1pt}
\setlength{\floatsep}{6pt plus 1pt minus 1pt}
\setlength{\intextsep}{6pt plus 1pt minus 1pt}

% ---------- Table I: Synthesis of literature ----------
\begin{table*}[!t]
\caption{Synthesis of the Twelve Reviewed Studies, Grouped by Theme}
\label{tab:summary}
\renewcommand{\arraystretch}{1.2}
\centering
\begin{tabular}{p{2.0cm}p{3.5cm}p{2.4cm}p{2.6cm}p{4.2cm}}
\toprule
\textbf{Theme} & \textbf{Study} & \textbf{Year / Method} & \textbf{Best Metric} & \textbf{Gap Left Open} \\
\midrule
\multirow{2}{*}{\parbox{2.0cm}{Systematic\\reviews}}
 & Porto~\cite{porto2024}             & 2024 / BERT, SVM, LSTM         & AUC $\sim$0.85+      & No regression; no SHAP \\
 & Multi-author SLR~\cite{slr2025}     & 2025 / RF, XGBoost, LR         & AUC 0.81--0.93       & Binary only; no public dataset \\
\midrule
\multirow{4}{*}{\parbox{2.0cm}{Empirical\\wait time / LOS}}
 & Wang (fairness)~\cite{wang2025fair} & 2025 / XGBoost + SHAP          & AUC $\approx$0.72    & Binary only; ESI-3 only \\
 & Ricciardi~\cite{ricciardi2024}      & 2024 / RF, DT, SVM             & AUC $\approx$0.80    & Binary; no SHAP; no segmentation \\
 & Seo~\cite{seo2024}                  & 2024 / LightGBM + NLP          & Best AUC-ROC overall & Single center; no fairness \\
 & Wu~\cite{wu2025}                    & 2025 / LightGBM (regression)   & Lowest MAE / RMSE    & Single center; no SHAP; no fairness \\
\midrule
\multirow{2}{*}{\parbox{2.0cm}{Benchmarks /\\interpretability}}
 & Xie~\cite{xie2022}                  & 2022 / GB, RF, MLP             & Beats MEWS/NEWS      & No wait time; no SHAP; no fairness \\
 & Sulaiman~\cite{sulaiman2025}        & 2025 / GB + TE2Rules           & Best classification  & Binary; single center; no regression \\
\midrule
\multirow{3}{*}{\parbox{2.0cm}{Fairness /\\overcrowding /\\advanced AI}}
 & Vural~\cite{vural2025}              & 2025 / TSiTPlus, XCMPlus       & MAE 4.19; R$^2$ 0.56 & Aggregate count; not patient-level \\
 & Lett~\cite{lett2025}                & 2025 / XGBoost (intersect.)    & AUROC 0.85           & Admission only; no regression \\
 & Sun (ED-Copilot)~\cite{sun2024}     & 2024 / BioGPT + RL             & Halved avg lab wait  & Binary targets; no ESI segmentation \\
\midrule
\multirow{1}{*}{\parbox{2.0cm}{Interpretable\\(directly compared)}}
 & Wang (interp.)~\cite{wang2025interp} & 2025 / XGB, RF, ANN, SVM      & Best Acc/Recall/F1   & Binary; ESI-3 only; no public data \\
\bottomrule
\end{tabular}
\end{table*}

% ---------- Table: Data dictionary, GROUPED BY CLINICAL ROLE ----------
\begin{table}[H]
\centering
\caption{Raw Data Dictionary, Grouped by Clinical Role ($N=9{,}216$)}
\label{tab:dictionary}
\renewcommand{\arraystretch}{1.15}
\begin{tabular}{p{1.9cm}p{3.0cm}p{2.5cm}}
\toprule
\textbf{Role} & \textbf{Field(s)} & \textbf{Type / Range} \\
\midrule
Identifier
 & \texttt{patient\_id} & int / unique \\
\midrule
\multirow{3}{*}{\parbox{1.9cm}{Demographics}}
 & \texttt{age}    & int / 0--100+ \\
 & \texttt{gender} & cat. / M, F \\
 & \texttt{race}   & cat. / 5 levels \\
\midrule
\multirow{2}{*}{\parbox{1.9cm}{Arrival\\context}}
 & \texttt{arrival\_hour}      & int / 0--23 \\
 & \texttt{arrival\_transport} & cat. / 4 levels \\
\midrule
\multirow{2}{*}{Triage}
 & \texttt{acuity}      & ord. / 1--5 (ESI) \\
 & \texttt{pain\_score} & float / 0--10 \\
\midrule
\multirow{6}{*}{\parbox{1.9cm}{Vital\\signs}}
 & \texttt{heartrate}   & float / bpm \\
 & \texttt{sbp}         & float / mmHg \\
 & \texttt{dbp}         & float / mmHg \\
 & \texttt{temperature} & float / $^\circ$F \\
 & \texttt{o2sat}       & float / \% \\
 & \texttt{resprate}    & float / br/min \\
\midrule
\multirow{2}{*}{Disposition}
 & \texttt{disposition}   & cat. / 5 levels \\
 & \texttt{was\_admitted} & bin / 0/1 \\
\midrule
Target
 & \texttt{ed\_los\_minutes} & float / minutes \\
\bottomrule
\end{tabular}
\end{table}

% ---------- Table: target distribution synthesis ----------
\begin{table}[H]
\centering
\caption{Target Variable Distribution, Grouped by Distribution Region}
\label{tab:target}
\renewcommand{\arraystretch}{1.15}
\begin{tabular}{p{2.0cm}lr}
\toprule
\textbf{Region} & \textbf{Statistic} & \textbf{Value (min)} \\
\midrule
\multirow{2}{*}{Centrality} & Mean   & 189.8 \\
                            & Median &  153.7 \\
\midrule
Spread                      & Std    & 141.7 \\
\midrule
\multirow{4}{*}{Tails}      & Min               &     8.6 \\
                            & 25th percentile   &    95.6 \\
                            & 75th percentile   &   241.6 \\
                            & Max               & 2{,}644.9 \\
\bottomrule
\end{tabular}
\end{table}

% ---------- Table: missing values, GROUPED BY MISSINGNESS SEVERITY ----------
\begin{table}[H]
\centering
\caption{Missing-Value Profile, Grouped by Severity (Total 5{,}156 cells, 3.3\%)}
\label{tab:missing}
\renewcommand{\arraystretch}{1.15}
\begin{tabular}{p{2.2cm}lrr}
\toprule
\textbf{Severity} & \textbf{Field} & \textbf{Count} & \textbf{\%} \\
\midrule
High ($\geq$10\%)
 & \texttt{temperature} & 1{,}105 & 11.99 \\
\midrule
\multirow{3}{*}{\parbox{2.2cm}{Moderate\\(8--10\%)}}
 & \texttt{resprate}    &  921 & 9.99 \\
 & \texttt{o2sat}       &  829 & 9.00 \\
 & \texttt{pain\_score} &  737 & 8.00 \\
\midrule
\multirow{3}{*}{\parbox{2.2cm}{Low\\($\leq$6\%)}}
 & \texttt{dbp}        & 552 & 5.99 \\
 & \texttt{sbp}        & 552 & 5.99 \\
 & \texttt{heartrate}  & 460 & 4.99 \\
\bottomrule
\end{tabular}
\end{table}

% ---------- Table: wrangling pipeline before/after, GROUPED by PROPERTY ----------
\begin{table}[H]
\centering
\caption{Wrangling Pipeline Before vs.\ After, Grouped by Property}
\label{tab:wrangling}
\renewcommand{\arraystretch}{1.15}
\begin{tabular}{p{2.0cm}lrr}
\toprule
\textbf{Property} & \textbf{Metric} & \textbf{Before} & \textbf{After} \\
\midrule
\multirow{2}{*}{Volume}
 & Rows                 & 9{,}216 & 9{,}216 \\
 & Columns              & 17       & 26       \\
\midrule
\multirow{2}{*}{Quality}
 & Missing cells        & 5{,}156 & 0        \\
 & Missing \%           &    3.3   & 0.0       \\
\midrule
\multirow{2}{*}{Schema}
 & Numeric features     & 13       & 22       \\
 & Categorical features &  4       &  5       \\
\midrule
\multirow{4}{*}{\parbox{2.0cm}{Target\\stability}}
 & Target mean (min)    & 189.8     & 189.8    \\
 & Target std  (min)    & 141.7     & 141.7    \\
 & Target min  (min)    &   8.6     &   8.6    \\
 & Target max  (min)    & 2{,}644.9 & 2{,}644.9 \\
\bottomrule
\end{tabular}
\end{table}

% ---------- Table: overall results synthesis (GROUPED BY MODEL FAMILY) ----------
\begin{table*}[!t]
\centering
\caption{Overall Model Performance, Grouped by Model Family
($n_{\mathrm{test}}=1{,}844$; CV on training fold)}
\label{tab:overall}
\renewcommand{\arraystretch}{1.2}
\begin{tabular}{p{2.0cm}lcccccc}
\toprule
\textbf{Family} & \textbf{Model} &
\textbf{Test MAE} & \textbf{Test RMSE} & \textbf{Test R$^2$} &
\textbf{CV MAE} & \textbf{CV RMSE} & \textbf{CV R$^2$} \\
\midrule
\multirow{2}{*}{\parbox{2.0cm}{Linear /\\regularized}}
 & Linear Regression & 89.10 & 130.35 & 0.141 & 90.01 & 131.33 & 0.145 \\
 & Ridge Regression  & \textbf{89.09} & \textbf{130.34} & \textbf{0.141} & \textbf{90.01} & \textbf{131.33} & \textbf{0.145} \\
\midrule
Bagged trees
 & Random Forest     & 91.03 & 131.96 & 0.120 & 92.23 & 133.58 & 0.114 \\
\midrule
\multirow{2}{*}{\parbox{2.0cm}{Boosted trees}}
 & XGBoost           & 93.80 & 134.91 & 0.080 & 94.81 & 137.18 & 0.065 \\
 & LightGBM          & 93.42 & 134.84 & 0.081 & 94.84 & 137.26 & 0.064 \\
\midrule
Sanity floor
 & Mean baseline     & 98.71 &   ---  &  ---  &  ---  &   ---  &  ---  \\
\bottomrule
\end{tabular}
\end{table*}

% ---------- Table: ESI-stratified, GROUPED BY ACUITY TIER ----------
\begin{table}[H]
\centering
\caption{ESI-Stratified Ridge Performance, Grouped by Acuity Tier}
\label{tab:esi}
\renewcommand{\arraystretch}{1.2}
\begin{tabular}{p{1.9cm}cccccc}
\toprule
\textbf{Tier} & \textbf{ESI} & \textbf{$N$} & \textbf{Mean LOS} & \textbf{MAE} & \textbf{RMSE} & \textbf{R$^2$} \\
\midrule
\multirow{2}{*}{High acuity}
 & 1 &   179 & 315.7 & 113.33 & 152.26 & $-$0.158 \\
 & 2 & 1{,}159 & 265.2 & 130.28 & 174.81 & $-$0.072 \\
\midrule
Mid acuity
 & 3 & 4{,}184 & 215.9 & 101.96 & 138.38 & $-$0.006 \\
\midrule
\multirow{2}{*}{Low acuity}
 & 4 & 2{,}706 & 145.0 &  69.96 & 101.62 & $-$0.015 \\
 & 5 &   988 &  91.4 &  36.96 &  48.25 & $-$0.066 \\
\bottomrule
\end{tabular}
\end{table}

% ---------- Table: SHAP global importance, GROUPED BY FEATURE FAMILY ----------
\begin{table}[H]
\centering
\caption{Global Mean $|$SHAP$|$ for LightGBM, Grouped by Feature Family}
\label{tab:shap}
\renewcommand{\arraystretch}{1.15}
\begin{tabular}{p{2.0cm}lr}
\toprule
\textbf{Family} & \textbf{Feature} & \textbf{Mean $|$SHAP$|$} \\
\midrule
Triage
 & \texttt{acuity}      & \textbf{46.07} \\
\midrule
\multirow{6}{*}{\parbox{2.0cm}{Vital signs}}
 & \texttt{o2sat}       &  7.15 \\
 & \texttt{dbp}         &  6.27 \\
 & \texttt{temperature} &  5.88 \\
 & \texttt{resprate}    &  5.20 \\
 & \texttt{heartrate}   &  4.96 \\
 & \texttt{sbp}         &  3.94 \\
\midrule
\multirow{2}{*}{\parbox{2.0cm}{Engineered\\physiology}}
 & \texttt{pulse\_pressure} & 5.85 \\
 & \texttt{shock\_index}    & 5.07 \\
\midrule
Pain
 & \texttt{pain\_score} &  3.70 \\
\midrule
\multirow{4}{*}{\parbox{2.0cm}{Demographics}}
 & \texttt{age}                & 5.87 \\
 & \texttt{race\_encoded}      & 3.49 \\
 & \texttt{gender\_encoded}    & 1.40 \\
 & \texttt{age\_group\_encoded}& 0.62 \\
\midrule
\multirow{5}{*}{\parbox{2.0cm}{Arrival /\\disposition}}
 & \texttt{arrival\_hour}             & 3.37 \\
 & \texttt{arrival\_transport\_enc}   & 2.15 \\
 & \texttt{is\_weekend}               & 1.32 \\
 & \texttt{was\_admitted}             & 1.14 \\
 & \texttt{arrival\_shift\_encoded}   & 0.96 \\
\bottomrule
\end{tabular}
\end{table}

% ---------- Table: fairness (race), GROUPED BY SUBGROUP SIZE ----------
\begin{table}[H]
\centering
\caption{Fairness Audit by Race/Ethnicity, Grouped by Subgroup Size}
\label{tab:fair_race}
\renewcommand{\arraystretch}{1.15}
\begin{tabular}{p{1.9cm}lrrr}
\toprule
\textbf{Tier} & \textbf{Subgroup} & \textbf{$N$} & \textbf{MAE} & \textbf{RMSE} \\
\midrule
\multirow{2}{*}{\parbox{1.9cm}{Large\\($N>300$)}}
 & WHITE                  & 1{,}032 & 90.78 & 135.47 \\
 & BLACK/AFR. AM.         &   366 & 91.23 & 130.00 \\
\midrule
\multirow{2}{*}{\parbox{1.9cm}{Medium\\(100--300)}}
 & HISPANIC               &   245 & 83.05 & 118.23 \\
 & ASIAN                  &   133 & 88.01 & 124.12 \\
\midrule
Small ($N<100$)
 & OTHER                  &    68 & 75.97 & 103.64 \\
\midrule
\multicolumn{3}{l}{\textbf{Range (max$-$min)}} & \textbf{15.26} & \textbf{31.83} \\
\bottomrule
\end{tabular}
\end{table}

% ---------- Table: fairness (gender), simple ----------
\begin{table}[H]
\centering
\caption{Fairness Audit by Gender (Ridge, Test Set)}
\label{tab:fair_gender}
\renewcommand{\arraystretch}{1.15}
\begin{tabular}{lrrr}
\toprule
\textbf{Subgroup} & \textbf{$N$} & \textbf{MAE} & \textbf{RMSE} \\
\midrule
Female (F) & 981 & 88.61 & 137.02 \\
Male (M)   & 863 & 89.64 & 122.31 \\
\midrule
\textbf{Range (max$-$min)} & --- & \textbf{1.03} & \textbf{14.71} \\
\bottomrule
\end{tabular}
\end{table}

% ---------- Table: fairness (age), GROUPED BY LIFE STAGE ----------
\begin{table}[H]
\centering
\caption{Fairness Audit by Age Group, Grouped by Life Stage}
\label{tab:fair_age}
\renewcommand{\arraystretch}{1.15}
\begin{tabular}{p{1.9cm}lrrr}
\toprule
\textbf{Stage} & \textbf{Subgroup} & \textbf{$N$} & \textbf{MAE} & \textbf{RMSE} \\
\midrule
Pediatric     & $\leq$18         & 148 & 95.60 & 150.62 \\
\midrule
\multirow{2}{*}{Adult}
              & 19--40 (Young)   & 600 & 93.34 & 136.35 \\
              & 41--65 (Middle)  & 901 & 84.10 & 125.58 \\
\midrule
Geriatric     & $>$65 (Elderly)  & 195 & 94.13 & 115.62 \\
\midrule
\multicolumn{3}{l}{\textbf{Range (max$-$min)}} & \textbf{11.50} & \textbf{35.00} \\
\bottomrule
\end{tabular}
\end{table}

\end{document}
